Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-15 mr-review.Rmd
added dagitty

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(magrittr)
```

@@

```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
knitr::include_graphics("figures/mr-gwas-counts.png")
```

@@

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. Because W affecs Y through pathways that don't involve exposure X, this scenario violates the exclusion restriction assumption."}

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(magrittr)
```

@@

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

ggplot2::ggsave("figures/mr-gwas-counts.png")

@@

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}

library(dagitty)
library(ggdag)
library(ggplot2)

d1 <- dagitty('dag {
bb="0,0,1,1"
G [pos="0,0"]
W [pos="0.5,0"]
X [exposure, pos="1.0,0"]
Y [outcome, pos="1.5,0"]
G -> W
W -> X
W -> Y [pos="1,1"]
X -> Y
}')

plot(d1)

```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which in turn affects Y. Unobserved confounder U affects both X and Y."}
d1 <- dagitty('dag {
bb="0,0,1,1"
G [pos="0,0"]
U [pos="0.75,10"]
X [exposure, pos="0.5,0"]
Y [outcome, pos="1.0,0"]
G -> X
U -> X
U -> Y
X -> Y
}'
)
plot(d1)
